Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-10 paper.tex
Updated introduction

% Problem definition
Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. Big~Brain, for
\todo{summarize sizes} instance, is a 3D histological image of the human brain that
represents 1~TB of raw data organized in 3600 planes at full resolution
and 76~GB at a 40-micrometer isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. This dataset
is being re-scanned at an even higher resolution, leading to 100~TB of
raw data. Other examples found in medical imaging, our primary domain
of interest, include high-resolution 3D electron microscopy (see,
e.g., ~\cite{bock2011network}) or micro- and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS~\cite{shvachko2010hadoop},
efficient software libraries are needed to split and merge them
efficiently. In particular, data reads and writes might result in
extensive seek times that drastically limit I/O performance.  In this
paper we introduce and compare a family of sequential algorithms to
split and merge images with minimal seek time.

Images are split into chunks representing 3D blocks or 2D slices,
which we assume fit in memory. A dataset such as Big~Brain would
perhaps be split into 125 chunks of 600~MB. The decision to split an
image into slices or blocks, and the size of the resulting chunks, is
usually done by the application. Some applications would commonly
require blocks, for instance spatial filtering, while other ones such
as acquisition artifact removal would rather work on
slices. Applications processing voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks. Flexibility is thus required in the splitting
scheme.

Seek time is highly dependent on the byte ordering in image files,
which we assume is arbitrary but known to the algorithm. The geometry
of the chunks to split or merge (slices or blocks) and their sizes are
also critical, which we again assume arbitrary but known to the
algorithm.  The main idea of our algorithms will be to convert data
orderings in memory before writing them to disk.

The literature on this problem is remarkably scarce. Parallel and
distributed image processing has obviously been extensively studied
and used in various
platforms~\cite{miguet1991elastic,tang2007eman2,yang2008parallel,braunl2013parallel,moise2013terabyte,bajcsy2013terabyte},
but methods have focused more on geometrical approaches to partition
images and on load-balancing and task scheduling techniques while we
aim at algorithms to efficiently split or merge images regardless of
the geometry of the chunks. Seek times are, however, often identified
as an issue but the preferred solution is usually to optimize data
storage formats for a certain application while we are searching for
algorithms that would reduce the seek time regardless of the data
format. For instance, the Open Connectome Data
Cluster~\cite{burns2013open} is a data warehouse system that allows
users to retrieve specific chunks of large image datasets. It
addresses seek times by designing a specific file format based on
space-filling curves, which elegantly preserves spatial proximity on
disk. Such specific data formats optimize the system for particular
chunk geometries, blocks in this case, while we aim at algorithms to
support multiple types of geometries.
%http://dl.acm.org/citation.cfm?id=218395 also minimizes seek time using a specific format.

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge
    terabyte-scale 3D images from blocks or slices.
  \item We model the performance (number of seeks) of these algorithms.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:experiments} details the results and
Sections~\ref{sec:discussion} discusses them.\todo{Revise when outline is final.}

@@

We consider a file format where voxels are written in column-major
order, that is, for an image of dimensions (i, j, k), i is the fastest
changing dimension, followed by dimension j, followed by dimension
k. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI (\url{https://nifti.nimh.nih.gov}) is such a format,
used by the vast majority of the neuroimaging community.

>>>>

% Problem definition
Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. Big~Brain, for
instance, is a 3D histological image of the
human brain that represents 1~TB of raw data organized in 3600 planes
at full resolution and 76~GB at a 40-micrometer isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Future
versions of this dataset will be acquired at an even higher resolution
and will represent 100~TB of raw data. Other examples found in medical
imaging, our primary domain of interest, include high-resolution 3D
electron microscopy (see, e.g., ~\cite{bock2011network}) or micro- and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as the Hadoop Distributed File System
(HDFS~\cite{shvachko2010hadoop}), software libraries are needed to
split and merge them efficiently, in particular to limit file
seeks. In this paper we introduce and compare a family of sequential
algorithms to split and merge images with limited numbers of seeks.

We assume that the high-resolution image is split into chunks
representing 3D blocks or 2D slices that fit in memory. A dataset such
as Big~Brain would perhaps be split into 125 chunks of 600~MB. The
decision to split an image into slices or blocks, and the size of the
chunks, is up to the application. Some applications, for instance spatial filtering, would
commonly require blocks while other
ones, such as acquisition artifact removal, would rather work on
slices. Applications that process voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks. Flexibility is thus required in the splitting
scheme.

We also assume that the byte organization in image files is arbitrary
but known to the algorithm. Some formats, for instance the file format
defined by the Neuroimaging Informatics Technology Initiative (NIfTI
-- \url{https://nifti.nimh.nih.gov}) store the complete image in
column-major order, that is, for an image of dimensions (i, j, k), i
is the fastest changing dimension, followed by dimension j, followed
by dimension k. Other formats, for instance HDF5-based MINC
2.0~\cite{vincent2016minc}, provide more flexibility by allowing data
to be partitioned in limited-sized chunks, each chunk being stored in
a specific order. Byte organization is obviously a critical factor of
seek time. The main idea of our algorithms will be to rearrange
ordering in memory before or after I/O operations.

The literature on this problem is remarkably scarce. Parallel and
distributed image processing has obviously been extensively studied
and used in various
platforms~\cite{miguet1991elastic,tang2007eman2,yang2008parallel,braunl2013parallel,moise2013terabyte,bajcsy2013terabyte},
but methods have focused on geometrical approaches to partition images
and on load-balancing and on task scheduling techniques. Instead, we
aim at algorithms to efficiently split or merge images regardless of
the geometry of the chunks. Although important seek times are often
identified as an issue, the preferred solution is usually to optimize
data storage formats for a certain application.  For instance, the
Open Connectome Data Cluster~\cite{burns2013open} is a data warehouse
system that allows users to retrieve specific 3D blocks from large
image datasets. It reduces seek times through a specific file format
based on space-filling curves, which elegantly preserves spatial
proximity on disk. On the contrary, we are searching for algorithms
that would reduce the seek time regardless of the data format, so that
applications with flexible splitting schemes can be served

%http://dl.acm.org/citation.cfm?id=218395 also minimizes seek time using a specific format.
To summarize, our paper makes the following contributions. (1) We
propose a set of algorithms to split and merge terabyte-scale 3D
images from 3D blocks or 2D slices. (2) We determine the complexity of
those algorithms in terms of numbers of seeks, as a function of the image size, splitting scheme and available memory. (3)We evaluate our algorithms using the Big~Brain dataset and two storage drives with different characteristics.

Section~\ref{sec:algos} presents our algorithms,
Section~\ref{sec:implementation} describes their implementation,
Section~\ref{sec:experiments} reports experimental results and
Section~\ref{sec:discussion} concludes the paper.

@@

We consider a file format where voxels are written in column-major
order. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI is such a format,
used by the vast majority of the neuroimaging community.
